The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-09-12 Independence_CH/discipline.tex
Appendix on lam_replacements. Macro for locales
Bye isatt, bye slanted

\appendix

\section{Discipline for relativization}

@@

Paulson only had, for each set, the relational version. It seemed
clearer to us to have a functional version of the relativized concept.
Going back to our example in \ref{sec:tools-relativization}, for the
concept $\isatt{cardinal}::\tyi \fun \tyi$ we want its relative
version
$\isatt{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
and the relational version of the latter
$\isatt{is{\uscore}cardinal}::(\tyi \fun \tyo) \fun \tyi \fun \tyi
\fun \tyo$.

Our first attempt of defining a discipline was inspired by
mathematical considerations: if we might prove that
$\isatt{is{\uscore}cardinal}$ is functional and also prove the
existence of a witness $\isatt{c}$ such that $\isatt{M(c)}$ and
$\isatt{is{\uscore}cardinal(M,x,c)}$ then
$\isatt{cardinal{\uscore}rel(M,x)}$ can be obtained by the operator of
definite descriptions.

Soon we realized that resorting to definite descriptions was needed
only for the most primitive concepts. In fact, once we have a
relativized concept, we can use it to define other relativizations.
For instance, $\isatt{cardinal{\uscore}rel}$ depends on having
relative versions of $\isatt{bij}$. Instead of relationalizing
$\isatt{bij}$ to get $\isatt{is{\uscore}bij}$ and then prove
uniqueness and existence of a witness, we define
$\isatt{bij{\uscore}rel}$ using $\isatt{inj{\uscore}rel}$ and
$\isatt{surj{\uscore}rel}$.

>>>>

\section{Discipline for relativization}
\label{sec:discipline-relativization}

@@

Paulson only had, for each set, the relational version. It seemed
clearer to us to have a functional version of the relativized concept.
Going back to our example in \ref{sec:tools-relativization}, for the
concept $\isa{cardinal}::\tyi \fun \tyi$ we want its relative
version
$\isa{cardinal{\uscore}rel}::(\tyi \fun \tyo) \fun \tyi \fun\tyi$
and the relational version of the latter
$\isa{is{\uscore}cardinal}::(\tyi \fun \tyo) \fun \tyi \fun \tyi
\fun \tyo$.

Our first attempt of defining a discipline was inspired by
mathematical considerations: if we might prove that
$\isa{is{\uscore}cardinal}$ is functional and also prove the
existence of a witness $\isa{c}$ such that $\isa{M(c)}$ and
$\isa{is{\uscore}cardinal(M,x,c)}$ then
$\isa{cardinal{\uscore}rel(M,x)}$ can be obtained by the operator of
definite descriptions.

Soon we realized that resorting to definite descriptions was needed
only for the most primitive concepts. In fact, once we have a
relativized concept, we can use it to define other relativizations.
For instance, $\isa{cardinal{\uscore}rel}$ depends on having
relative versions of $\isa{bij}$. Instead of relationalizing
$\isa{bij}$ to get $\isa{is{\uscore}bij}$ and then prove
uniqueness and existence of a witness, we define
$\isa{bij{\uscore}rel}$ using $\isa{inj{\uscore}rel}$ and
$\isa{surj{\uscore}rel}$.

### 2022-09-12 Independence_CH/forcing.tex
Bye isatt, bye slanted

states that Zermelo set theory holds in $M$. The Union Axiom (\isatt{Union{\uscore}ax}), for

@@

\forall x[\isatt{\#\#}M].\ \exists z[\isatt{\#\#}M].\ \isatt{big{\uscore}union}(\isatt{\#\#}M,x,z)

@@

  \isatt{Union{\uscore}ax}(\isatt{\#\#}M) \lsii M, [\,]\models \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}}

@@

where \isatt{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the

@@

  \isatt{separation}(N,Q)

@@

  \isatt{strong{\uscore}replacement}(N,R)

@@

$\isatt{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes

@@

\varphi \in \formula  \limp \mathit{env} \in \isatt{list}(M) \limp \isatt{arity}(\varphi) \leq 2+ \isatt{length}(\mathit{env}) \limp \\
 \isatt{strong{\uscore}replacement}(\isatt{\#\#} M, \lambda x\, y.\ (M, [x,y]

@@

In turn, the \isatt{{\isasymcdot}Replacement{\isasymcdot}} function

@@

In \session{ZF-Constructible} some results are proved assuming some instances of
separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
our locale \isatt{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
es distinto a Reemplazo y se nombra después)

@@

\sout{The locales gathering all 22 replacement instances necessary are named
\locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF3},
\locale{M{\uscore}ZF{\uscore}ground},
\locale{M{\uscore}ZF{\uscore}ground{\uscore}notCH}, and
\locale{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
Section~\ref{sec:repl-instances} for details).}
These assumptions are then invoked to interpret at the class
$\isatt{\#\#} M$ the relevant locales appearing in
\session{ZF-Constructible}, and further ones required for the relative
results from Section~\ref{sec:extension-isabellezf}. See
Section~\ref{sec:repl-instances} for details.

@@

\subsection{The fundamental theorems}\label{sec:fundamental-theorems}
At this point, we work inside the locale \isatt{M{\uscore}ctm1} that
assumes $M$ to be
countable and transitive, and satisfies some fragment of
$\ZFC$%
\footnote{%
Namely, Zermelo set theory plus the 7 replacement instances included in
the locales \locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
}.
This is further extended by assuming a forcing notion
$\lb \PP, {\preceq} ,\1\rb \in M$. The actual implementation reads:

@@

\begin{theorem}\label{th:forcing-thms}
  For every
  $\phi\in\formula$ with $\isatt{arity}(\phi)\leq n$ and $\tau_1,\dots,\tau_n\in M$,
  \begin{enumerate}
  \item\label{item:definability} (Definability)
    $\forceisa(\phi)\in\formula$, 
  \end{enumerate}
  where the 
  arity of $\forceisa(\phi)$ is at most $\isatt{arity}(\phi) + 4$; and if
  “$p \forces \phi\ [\tau_1,\dots,\tau_n]$”
  denotes
  “$M, [p,\PP,\preceq,\1, \tau_1,\dots,\tau_n]  \models
  \forceisa(\phi)$”, we have:
  \begin{enumerate}
    \setcounter{enumi}{1}
  \item\label{item:truth-lemma} (Truth Lemma) for every $M$-generic $G$,
    \[
      \exists p\in G.\ \; p \forces \phi\ [\tau_1,\dots,\tau_n]
    \]
    is equivalent to 
    \[
      M[G], [\val(G,\tau_1),\dots,\val(G,\tau_n)]
      \models \phi.
    \]
  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\tau_1,\dots,\tau_n]$
    if and only if 
    $\{q\in \PP :  q \forces \phi\ [\tau_1,\dots,\tau_n]\}$
    is dense below $p$.
  \end{enumerate}
\end{theorem}
The items in Theorem~\ref{th:forcing-thms} appear as three
separated lemmas in the theory
\theory{Forcing{\uscore}Theorems} of our
\session{Independence\_CH} session \cite{Independence_CH-AFP}.
For instance, the Truth Lemma is stated as
follows:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ truth{\isacharunderscore}{\kern0pt}lemma{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymle}length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
\end{isabelle}
%% where the $\isatt{map}$ function that applies a function to
%% each element of a list and
where the $\forces$ notation (and its precedence) had already been set up in the
\theory{Forces{\uscore}Definition} theory.

Kunen first describes forcing for atomic formulas using a mutual
recursion
%% \begin{multline*}
%%   \forceseq (p,t_1,t_2) \defi 
%%   \forall s\in\dom(t_1)\cup\dom(t_2).\ \forall q\pleq p .\\
%%   \forcesmem(q,s,t_1)\lsii \forcesmem(q,s,t_2)
%% \end{multline*}
%% \begin{multline*}
%%   \forcesmem(p,t_1,t_2) \defi  \forall v\pleq p. \ \exists q\pleq v.\\  
%%   \exists s.\ \exists r\in \PP .\ \lb s,r\rb \in  t_2 \land q
%%   \pleq r \land \forceseq(q,t_1,s)
%% \end{multline*}
but then \cite[p.~257]{kunen2011set} it is cast as a single
recursively defined function $F$ over a well-founded  relation $R$.
In our formalization, these are called $\frcat$ and 
$\isatt{frecR}$, respectively, and are defined on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
$\mathit{ft}\in\{0,1\}$ indicates whether the atomic formula being
forced is an equality or a membership, respectively).
Forcing for general formulas is then defined by recursion on the
datatype $\formula$ as indicated above. Technical details on the
implementation and proofs of the
Forcing Theorems have been spelled out in our
\cite{2020arXiv200109715G}.

>>>>

states that Zermelo set theory holds in $M$. The Union Axiom (\isa{Union{\uscore}ax}), for

@@

\forall x[\isa{\#\#}M].\ \exists z[\isa{\#\#}M].\ \isa{big{\uscore}union}(\isa{\#\#}M,x,z)

@@

  \isa{Union{\uscore}ax}(\isa{\#\#}M) \lsii M, [\,]\models \isa{{\isasymcdot}Union\ Ax{\isasymcdot}}

@@

where \isa{{\isasymcdot}Union\ Ax{\isasymcdot}} is the $\formula$ code for the

@@

  \isa{separation}(N,Q)

@@

  \isa{strong{\uscore}replacement}(N,R)

@@

$\isa{replacement{\uscore}assm}(M,\mathit{env},\phi)$ predicate ($@$ denotes

@@

\varphi \in \formula  \limp \mathit{env} \in \isa{list}(M) \limp \isa{arity}(\varphi) \leq 2+ \isa{length}(\mathit{env}) \limp \\
 \isa{strong{\uscore}replacement}(\isa{\#\#} M, \lambda x\, y.\ (M, [x,y]

@@

In turn, the \isa{{\isasymcdot}Replacement{\isasymcdot}} function

@@

In \session{ZF-Constructible} some results are proved assuming some instances of
separation or replacement (mejor si ponemos ejemplo concreto). In contrast,
our locale \isa{M{\uscore}Z{\uscore}basic} assumes the full scheme of separation in terms
of satisfaction of some internalized formula but no instance of replacement. (Mostrar el enunciado de separation\_ax:
es distinto a Reemplazo y se nombra después)

@@

\sout{The locales gathering all 22 replacement instances necessary are named
\locale{M{\uscore}ZF1} through \locale{M{\uscore}ZF3},
\locale{M{\uscore}ZF{\uscore}ground},
\locale{M{\uscore}ZF{\uscore}ground{\uscore}notCH}, and
\locale{M{\uscore}ZF{\uscore}ground{\uscore}CH} (see
Section~\ref{sec:repl-instances} for details).}
These assumptions are then invoked to interpret at the class
$\isa{\#\#} M$ the relevant locales appearing in
\session{ZF-Constructible}, and further ones required for the relative
results from Section~\ref{sec:extension-isabellezf}. See
Section~\ref{sec:repl-instances} for details.

@@

\subsection{The fundamental theorems}\label{sec:fundamental-theorems}
At this point, we work inside the locale \isa{M{\uscore}ctm1} that
assumes $M$ to be
countable and transitive, and satisfies some fragment of
$\ZFC$%
\footnote{%
Namely, Zermelo set theory plus the 7 replacement instances included in
the locales \locale{M{\uscore}ZF1} and \locale{M{\uscore}ZF{\uscore}ground}.
}.
This is further extended by assuming a forcing notion
$\lb \PP, {\preceq} ,\1\rb \in M$. The actual implementation reads:

@@

\begin{theorem}\label{th:forcing-thms}
  For every
  $\phi\in\formula$ with $\isa{arity}(\phi)\leq n$ and $\tau_1,\dots,\tau_n\in M$,
  \begin{enumerate}
  \item\label{item:definability} (Definability)
    $\forceisa(\phi)\in\formula$, 
  \end{enumerate}
  where the 
  arity of $\forceisa(\phi)$ is at most $\isa{arity}(\phi) + 4$; and if
  “$p \forces \phi\ [\tau_1,\dots,\tau_n]$”
  denotes
  “$M, [p,\PP,\preceq,\1, \tau_1,\dots,\tau_n]  \models
  \forceisa(\phi)$”, we have:
  \begin{enumerate}
    \setcounter{enumi}{1}
  \item\label{item:truth-lemma} (Truth Lemma) for every $M$-generic $G$,
    \[
      \exists p\in G.\ \; p \forces \phi\ [\tau_1,\dots,\tau_n]
    \]
    is equivalent to 
    \[
      M[G], [\val(G,\tau_1),\dots,\val(G,\tau_n)]
      \models \phi.
    \]
  \item \label{item:density-lemma} (Density Lemma) $p \forces \phi\ [\tau_1,\dots,\tau_n]$
    if and only if 
    $\{q\in \PP :  q \forces \phi\ [\tau_1,\dots,\tau_n]\}$
    is dense below $p$.
  \end{enumerate}
\end{theorem}
The items in Theorem~\ref{th:forcing-thms} appear as three
separated lemmas in the theory
\theory{Forcing{\uscore}Theorems} of our
\session{Independence\_CH} session \cite{Independence_CH-AFP}.
For instance, the Truth Lemma is stated as
follows:
\begin{isabelle}
\isacommand{lemma}\isamarkupfalse%
\ truth{\isacharunderscore}{\kern0pt}lemma{\isacharcolon}{\kern0pt}\isanewline
\ \ \isakeyword{assumes}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isasymphi}{\isasymin}formula{\isachardoublequoteclose}\isanewline
\ \ \ \ {\isachardoublequoteopen}env{\isasymin}list{\isacharparenleft}{\kern0pt}M{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\ {\isachardoublequoteopen}arity{\isacharparenleft}{\kern0pt}{\isasymphi}{\isacharparenright}{\kern0pt}{\isasymle}length{\isacharparenleft}{\kern0pt}env{\isacharparenright}{\kern0pt}{\isachardoublequoteclose}\isanewline
\ \ \isakeyword{shows}\isanewline
\ \ \ \ {\isachardoublequoteopen}{\isacharparenleft}{\kern0pt}{\isasymexists}p{\isasymin}G{\isachardot}{\kern0pt}\ p\ {\isasymtturnstile}\ {\isasymphi}\ env{\isacharparenright}{\kern0pt}\ \ \ {\isasymlongleftrightarrow}\ \ \ M{\isacharbrackleft}{\kern0pt}G{\isacharbrackright}{\kern0pt}{\isacharcomma}{\kern0pt}\ map{\isacharparenleft}{\kern0pt}val{\isacharparenleft}{\kern0pt}G{\isacharparenright}{\kern0pt}{\isacharcomma}{\kern0pt}env{\isacharparenright}{\kern0pt}\ {\isasymTurnstile}\ {\isasymphi}{\isachardoublequoteclose}
\end{isabelle}
%% where the $\isa{map}$ function that applies a function to
%% each element of a list and
where the $\forces$ notation (and its precedence) had already been set up in the
\theory{Forces{\uscore}Definition} theory.

Kunen first describes forcing for atomic formulas using a mutual
recursion
%% \begin{multline*}
%%   \forceseq (p,t_1,t_2) \defi 
%%   \forall s\in\dom(t_1)\cup\dom(t_2).\ \forall q\pleq p .\\
%%   \forcesmem(q,s,t_1)\lsii \forcesmem(q,s,t_2)
%% \end{multline*}
%% \begin{multline*}
%%   \forcesmem(p,t_1,t_2) \defi  \forall v\pleq p. \ \exists q\pleq v.\\  
%%   \exists s.\ \exists r\in \PP .\ \lb s,r\rb \in  t_2 \land q
%%   \pleq r \land \forceseq(q,t_1,s)
%% \end{multline*}
but then \cite[p.~257]{kunen2011set} it is cast as a single
recursively defined function $F$ over a well-founded  relation $R$.
In our formalization, these are called $\frcat$ and 
$\isa{frecR}$, respectively, and are defined on tuples $\lb \mathit{ft},t_1,t_2,p\rb$ (where
$\mathit{ft}\in\{0,1\}$ indicates whether the atomic formula being
forced is an equality or a membership, respectively).
Forcing for general formulas is then defined by recursion on the
datatype $\formula$ as indicated above. Technical details on the
implementation and proofs of the
Forcing Theorems have been spelled out in our
\cite{2020arXiv200109715G}.

### 2022-09-12 Independence_CH/lessons.tex
Signaling edits after meeting with Miguel. Proposed new section on replacement instances, and new names in appendix
Bye isatt, bye slanted

As a result from this, some design choices that seemed reasonable at
first were proved to be inconvenient. For instance, we should had
better used predicates (of type $\tyi\fun\tyi\fun\tyo$) for the
forcing posets' order relations; this is the way they
are presented in the \session{Delta\_System\_Lemma} session. A similar
problem is that we require the forcing poset to be an element of $M$,
so the present infrastructure does not allow class forcing out of the
box. (The latter change seems to be rather straightforward, but the
former does not.)

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families;
  thus limiting us to $\kappa=\aleph_1$ case of the $\kappa$-chain condition, and dodging the relativization
  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $(<\omega+1)$-closed, that is, under plain $\omega$-sequences).
\end{itemize}
In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isatt{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

@@

Nevertheless, the experience of the CS members of the team (and
community) is invaluable for the kind of engineering that is needed in
a large project like the present one. For example, it is
(mathematically) misleading when automatic tools (\isatt{simp},
\isatt{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeeds but with 8
variables doesn't). Scale issues are very easily disregarded in the
abstract but, as an example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle2021-1 (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.

>>>>

As a result from this, some design choices that seemed reasonable at
first were proved to be inconvenient. For instance, we should had
better used predicates (of type $\tyi\fun\tyi\fun\tyo$) for the
forcing posets' order relations; this is the way they
are presented in the \session{Delta\_System\_Lemma} session. A similar
problem is that we require the forcing poset to be an element of $M$,
so the present infrastructure does not allow class forcing out of the
box. (The latter change seems to be rather straightforward, but the
former does not.)\\
\fbox{Note that this is a consequence of following
  Kunen's blueprint for the formalization}

Nearly the final stage of the project, we decided to go for the minimal
set of definitions and versions of lemmas that were needed to obtain
our target results. For example,
\begin{itemize}
\item
  we only proved the Delta System Lemma for $\aleph_1$-sized families;
  thus limiting us to $\kappa=\aleph_1$ case of the $\kappa$-chain condition, and dodging the relativization
  of the material on cofinalities \cite{Delta_System_Lemma-AFP};
\item we showed preservation of sequences by considering countably
  closed forcings (in fact, we formalized the bare minimum requirement
  of being $(<\omega+1)$-closed, that is, under plain $\omega$-sequences).
\end{itemize}
In doing this we went against the tried and
true advice that one should formalize the most general version of the
results available. Another shortcut we took was to
simplify some proofs by appealing to the countability of the ground
model; this is the case of
\isa{definition{\uscore}of{\uscore}forces} and the result on forcing
values of a function.

@@

Nevertheless, the experience of the CS members of the team (and
community) is invaluable for the kind of engineering that is needed in
a large project like the present one. For example, it is
(mathematically) misleading when automatic tools (\isa{simp},
\isa{auto}, etc) stop working just because of the sheer size of the
goal (v.g., the same statement with 7 variables succeeds but with 8
variables doesn't). Scale issues are very easily disregarded in the
abstract but, as an example, the formula $\forceisa({\cdot}0\in 1{\cdot})$
can be explicitly printed by Isabelle2021-1 (it spans nearly 20k symbols), but
$\forceisa({\cdot}\neg{\cdot}\neg{\cdot}0\in 1{\cdot}{\cdot}{\cdot})$
can not.

### 2022-09-12 Independence_CH/main_defs.tex
Bye isatt, bye slanted

The reader trusting
all the libraries on which our development is based, might jump
directly to Section~\ref{sec:relative-arith}, which treats relative
cardinal arithmetic as implemented in
\isatt{T{\kern0pt}r{\kern0pt}a{\kern0pt}n{\kern0pt}s{\kern0pt}i{\kern0pt}t{\kern0pt}i{\kern0pt}v{\kern0pt}e{\kern0pt}{\char`\_}{\kern0pt}M{\kern0pt}o{\kern0pt}d{\kern0pt}e{\kern0pt}l{\kern0pt}s{\kern0pt}}. But in case one wants to dive deeper, the
following sections treat some basic concepts of the ZF logic
(Section~\ref{sec:def-main-ZF}) and in the
\session{ZF-Constructible} library (Section~\ref{sec:def-main-relative})
on which our definitions are built.

@@

\begin{isabelle}
\isacommand{lemma}
\ empty{\uscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isasymforall}x{\isachardot}{\kern0pt}\ x\ {\isasymnotin}\ {\isadigit{0}}{\isachardoublequoteclose}%
\isanewline
\isacommand{lemma}
\ succ{\uscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}succ{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ i\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}i{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}%
\end{isabelle}
%
we can define the set of natural numbers \isa{{\isasymomega}}. In the
sources, it is  defined as a fixpoint, but here we just write
its characterization as the first limit ordinal.%
\begin{isabelle}%
Ord{\isacharparenleft}{\kern0pt}{\isasymomega}{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isadigit{0}}\ {\isacharless}{\kern0pt}\ {\isasymomega}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isachardot}{\kern0pt}\ y\ {\isacharless}{\kern0pt}\ {\isasymomega}\ {\isasymlongrightarrow}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\ {\isacharless}{\kern0pt}\ {\isasymomega}{\isacharparenright}{\kern0pt}\isasep\isanewline%
Ord{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isadigit{0}}\ {\isacharless}{\kern0pt}\ i\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isachardot}{\kern0pt}\ y\ {\isacharless}{\kern0pt}\ i\ {\isasymlongrightarrow}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\ {\isacharless}{\kern0pt}\ i{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isasymomega}\ {\isasymle}\ i%
\end{isabelle}%
Then, addition and predecessor on \isa{{\isasymomega}} are inductively
characterized as follows:%
\begin{isabelle}%
m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ succ{\isacharparenleft}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ n{\isacharparenright}{\kern0pt}\isasep\isanewline%
m\ {\isasymin}\ {\isasymomega}\ {\isasymLongrightarrow}\ m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ {\isadigit{0}}\ {\isacharequal}{\kern0pt}\ m\isasep\isanewline\isanewline%
pred{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{0}}\isasep\isanewline%
pred{\isacharparenleft}{\kern0pt}succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ y%
\end{isabelle}%
Lists on a set \isa{A} can be characterized by being
recursively generated from the empty list \isa{{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}} and the
operation \isa{Cons} that adds a new element to the left end;
the induction theorem for them shows that the characterization is
“complete”. (Mind the
\isatt{\isasymlbrakk P; Q\isasymrbrakk\ \isasymLongrightarrow\ R}
abbreviation for
\isatt{P\ \isasymLongrightarrow\ Q\ \isasymLongrightarrow\ R}.)

@@

The \isatt{arity} function strictly bounding the free de Bruijn

@@

actually is a sentence, it does not depend on \isatt{env}, which

>>>>

The reader trusting
all the libraries on which our development is based, might jump
directly to Section~\ref{sec:relative-arith}, which treats relative
cardinal arithmetic as implemented in
\isa{T{\kern0pt}r{\kern0pt}a{\kern0pt}n{\kern0pt}s{\kern0pt}i{\kern0pt}t{\kern0pt}i{\kern0pt}v{\kern0pt}e{\kern0pt}{\char`\_}{\kern0pt}M{\kern0pt}o{\kern0pt}d{\kern0pt}e{\kern0pt}l{\kern0pt}s{\kern0pt}}. But in case one wants to dive deeper, the
following sections treat some basic concepts of the ZF logic
(Section~\ref{sec:def-main-ZF}) and in the
\session{ZF-Constructible} library (Section~\ref{sec:def-main-relative})
on which our definitions are built.

@@

\begin{isabelle}
\isacommand{lemma}
\ empty{\uscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}{\isasymforall}x{\isachardot}{\kern0pt}\ x\ {\isasymnotin}\ {\isadigit{0}}{\isachardoublequoteclose}%
\isanewline
\isacommand{lemma}
\ succ{\uscore}{\kern0pt}def{\isacharprime}{\kern0pt}{\isacharcolon}{\kern0pt}\ {\isachardoublequoteopen}succ{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ i\ {\isasymunion}\ {\isacharbraceleft}{\kern0pt}i{\isacharbraceright}{\kern0pt}{\isachardoublequoteclose}%
\end{isabelle}
%
we can define the set of natural numbers \isa{{\isasymomega}}. In the
sources, it is  defined as a fixpoint, but here we just write
its characterization as the first limit ordinal.%
\begin{isabelle}%
Ord{\isacharparenleft}{\kern0pt}{\isasymomega}{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isadigit{0}}\ {\isacharless}{\kern0pt}\ {\isasymomega}\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isachardot}{\kern0pt}\ y\ {\isacharless}{\kern0pt}\ {\isasymomega}\ {\isasymlongrightarrow}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\ {\isacharless}{\kern0pt}\ {\isasymomega}{\isacharparenright}{\kern0pt}\isasep\isanewline%
Ord{\isacharparenleft}{\kern0pt}i{\isacharparenright}{\kern0pt}\ {\isasymand}\ {\isadigit{0}}\ {\isacharless}{\kern0pt}\ i\ {\isasymand}\ {\isacharparenleft}{\kern0pt}{\isasymforall}y{\isachardot}{\kern0pt}\ y\ {\isacharless}{\kern0pt}\ i\ {\isasymlongrightarrow}\ succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}\ {\isacharless}{\kern0pt}\ i{\isacharparenright}{\kern0pt}\ {\isasymLongrightarrow}\ {\isasymomega}\ {\isasymle}\ i%
\end{isabelle}%
Then, addition and predecessor on \isa{{\isasymomega}} are inductively
characterized as follows:%
\begin{isabelle}%
m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ succ{\isacharparenleft}{\kern0pt}n{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ succ{\isacharparenleft}{\kern0pt}m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ n{\isacharparenright}{\kern0pt}\isasep\isanewline%
m\ {\isasymin}\ {\isasymomega}\ {\isasymLongrightarrow}\ m\ {\isacharplus}{\kern0pt}\isactrlsub {\isasymomega}\ {\isadigit{0}}\ {\isacharequal}{\kern0pt}\ m\isasep\isanewline\isanewline%
pred{\isacharparenleft}{\kern0pt}{\isadigit{0}}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ {\isadigit{0}}\isasep\isanewline%
pred{\isacharparenleft}{\kern0pt}succ{\isacharparenleft}{\kern0pt}y{\isacharparenright}{\kern0pt}{\isacharparenright}{\kern0pt}\ {\isacharequal}{\kern0pt}\ y%
\end{isabelle}%
Lists on a set \isa{A} can be characterized by being
recursively generated from the empty list \isa{{\isacharbrackleft}{\kern0pt}{\isacharbrackright}{\kern0pt}} and the
operation \isa{Cons} that adds a new element to the left end;
the induction theorem for them shows that the characterization is
“complete”. (Mind the
\isa{\isasymlbrakk P; Q\isasymrbrakk\ \isasymLongrightarrow\ R}
abbreviation for
\isa{P\ \isasymLongrightarrow\ Q\ \isasymLongrightarrow\ R}.)

@@

The \isa{arity} function strictly bounding the free de Bruijn

@@

actually is a sentence, it does not depend on \isa{env}, which
